A Linear Solution of the Linearity Challenge

### 2025-12-22 paper/introduction.tex
drafted intro

This paper presents a solution to the linearity challenge.
%
The calculus that we formalize, dubbed \Calculus, and the one in the challenge
description differ for two key aspects.
%
First of all, the challenge only considers a minimal calculus of first-order
sessions, whereas \LCP is a fully-featured calculus that includes support for
higher-order, polymorphic, linear and shared channels. Also, \LCP is closely
related to \CP~\citep{Wadler14} and its type system coincides with the proof
system of classical linear logic.
%
The second difference is that the calculus in the challenge is equipped with
\emph{native} sessions, whereas \LCP features shared channels, which are
unrestricted, and linear channels, which are meant to be used for a
\emph{single} communication. Structured sessions involving the exchange of
several messages can be modeled using \emph{explicit continuations} according to
the well-known encoding of (binary) sessions into linear
channels~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
%
Overall, \LCP has the same expressive power of \CP, but is a lower-level
calculus in which sessions are encodable instead of being featured natively.

Compared to the other analogous formalizations of session/linear
calculi~\cite{GotoEtAl16,Thiemann19,CicconePadovani20,CastroFerreiraYoshida20,RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,JacobsBalzerKrebbers22,SanoKavanaghPientka23,ZackonEtAl25},
the one we present here is both the most complete (in terms of features
supported by the calculus) and the most streamlined (in terms of code size). It
is also one of only two formalizations that prove the deadlock freedom property
for a session calculus, and it does so with substantially less code than the
existing formalization~\cite{JacobsBalzerKrebbers22}.

We acknowledge that \LCP is not

The differences between \LCP and the calculus are functional to the simplicity
of its formalization.

the outcome of our investigation are somehow surprising in the sense that the
premises of the challenge and the assumption that the main technical challenge
is the representation of contexts

congetturiamo che la semplicità della formalizzazione derivi dalle
caratteristiche del calcolo più che dalle tecniche utilizzate, che alla fine
sono piuttosto standard.

We value these aspects of \Calculus because their combined effect allows us to
obtain a remarkably simple formalization.
%
The fact of working with a calculus based on linear logic prevents \emph{by
construction} the same (sequential) process to own both endpoints of a session.
This situation is clearly undesirable since it does not correspond to a useful
pattern of interaction (every meaningful session requires its endpoints to be
used by parallel processes) and is a potential source of deadlocks.
%
From the standpoint of the formalization, where the representation of channel
names is a primary design choice, it spares us the need to \emph{distinguish}
the two endpoints of a session, \eg by means of polarities~\citep{GayHole05} or
by using different names connected by the same
binder~\citep{Vasconcelos12,GayVasconcelos25}.
%
Also, the fact of working with a calculus with \emph{linear} channels and
explicit continuations spares us the need to \emph{update} the type of channels
in typing contexts. When a channel is used it is effectively consumed, therefore
its type can be \emph{removed} from the typing context and the type of the
continuation channel is \emph{added} back to the typing context. As is turns
out, removing and adding types is easier than updating them. Even more so
considering that the type of continuation channels can be conveniently added
\emph{at the beginning} of a typing context, since such channels are fresh by
definition.

BISOGNA PARLARE DI RELATED WORK

The rest of the paper is structured as follows. \Cref{sec:calculus} describes
the syntax and the operational semantics of \Calculus and states the properties
of well-typed processes that we formalize and prove. \Cref{sec:agda} outlines
the essential aspects of the formalization with particular emphasis on the
representation of channel names and typing contexts. \Cref{sec:conclusion}
summarizes our contributions, discusses some extensions and future work. The
formalization has been carried out using Agda and its full code is available
from a public\Luca{Related work? Confronto con altre formalizzazioni?}
repository.\footnote{\url{https://github.com/boystrange/LinearityChallenge/tree/main/cp}}

>>>>

This paper proposes a logical and linear solution of the linearity challenge. By
``logical and linear solution'' we mean that the calculus that we formalize,
dubbed \Calculus, and the one in the challenge description differ for two key
aspects.
%
First of all, the challenge considers an \emph{ad hoc}, minimal calculus of
first-order sessions, whereas \LCP is a fully-featured calculus that supports
higher-order, polymorphic, linear and shared channels. Also, \LCP is closely
related to \CP~\citep{Wadler14} and its type system coincides with the proof
system of classical linear logic.
%
The second difference is that the calculus in the challenge is equipped with
\emph{native} sessions, whereas \LCP features linear channels that are meant to
be used for a \emph{single} communication. Structured sessions involving the
exchange of several messages can be modeled using \emph{explicit continuations}
according to the well-known encoding of (binary) sessions into linear
channels~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
%
Overall, \LCP is a lower-level calculus in which sessions are encodable instead
of being featured natively, but it has the same expressive power of \CP.

We value these differences between \LCP and the calculus in the challenge
description because we think that they contribute to the simplicity of the
formalization more than the adopted formalization techniques, which end up being
fairly standard.\Luca{Our work fits the spirit of the challenge identifying not
only the most appropriate techniques but also the most appropriate models that
aid the formalization of concurrent/distributed models of computation.}
%
The fact of working with a calculus based on linear logic prevents \emph{by
construction} the same (sequential) process to own both endpoints of a session.
This situation is clearly undesirable since it does not correspond to a useful
pattern of interaction (every meaningful session requires its endpoints to be
used by parallel processes) and is a potential source of deadlocks.
%
From the standpoint of the formalization, where the representation of channel
names is a primary design choice, it spares us the need to \emph{distinguish}
the two endpoints of a session, \eg by means of polarities~\citep{GayHole05} or
by using different names connected by the same
binder~\citep{Vasconcelos12,GayVasconcelos25}.
%
Also, the fact of working with a calculus with \emph{linear} channels and
explicit continuations spares us the need to \emph{update} the type of channels
in typing contexts. When a channel is used it is effectively consumed, therefore
its type can be \emph{removed} from the typing context and the type of the
continuation channel is \emph{added} back to the typing context. As is turns
out, removing and adding types is easier than updating them. Even more so
considering that the type of continuation channels can be conveniently added
\emph{at the beginning} of a typing context, since such channels are fresh by
definition.

% Compared to the other analogous formalizations of session/linear
% calculi~\cite{GotoEtAl16,Thiemann19,CicconePadovani20,CastroFerreiraYoshida20,RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,JacobsBalzerKrebbers22,SanoKavanaghPientka23,ZackonEtAl25},
% the one we present here is both the most complete (in terms of features
% supported by the calculus) and the most streamlined (in terms of code size). It
% is also one of only two formalizations that prove the deadlock freedom property
% for a session calculus, and it does so with substantially less code than the
% existing formalization~\cite{JacobsBalzerKrebbers22}.

\subparagraph*{Structure of the paper.}
%
\Cref{sec:calculus} describes the syntax and the operational semantics of \LCP
and states the properties of well-typed processes that we formalize and prove,
namely typing preservation, deadlock freedom and runtime safety.
%
\Cref{sec:agda} presents a roadmap of the Agda formalization and discusses its
key aspects with particular emphasis on the representation of channel names and
of typing contexts.
%
\Cref{sec:examples} illustrates a few examples of \LCP processes as they are
represented in Agda and shows how to use Agda as an interpreter for \LCP.
%
\Cref{sec:related} discusses related work more in detail by providing a
qualitative and quantitative comparison between the known formalizations of
session/linear calculi, including the one for \LCP.
%
\Cref{sec:conclusion} summarizes our contributions and discusses some ongoing
and future work.

The formalization has been carried out using Agda and the source code is
available on Github at
\url{https://github.com/boystrange/LinearityChallenge/tree/main/lcp}.
